---
title: "PADP 7120 Data Applications in PA"
subtitle: "Regression with Interactions"
author: "Alex Combs"
institute: "UGA | SPIA | PADP"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  xaringan::moon_reader:
    css: 'mypres.css'
    nature:
      beforeInit: "macros.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
# Outline

```{r, include=FALSE}
library(tidyverse)
library(moderndive)
library(knitr)
library(carData)
options(scipen = 999)
```

- Interact numerical and categorical variables

---
# Parallel Slopes

```{r, echo=FALSE, fig.width=8, fig.align='center'}
ggplot(Salaries, aes(x = yrs.service, y = salary, color = sex)) +
  geom_point(size = 4, alpha = 0.5) +
  geom_abline(slope = 747, intercept = 92357, color = 'red', linewidth = 2) +
  geom_abline(slope = 747, intercept = 101429, color = 'blue', linewidth = 2) +
  theme_minimal() +
  theme(text = element_text(size = 20))
```

---
# Example: Parallel Slopes

$$Salary=\beta_0+\beta_1YrsEmployed+\beta_2Sex + \epsilon$$
```{r}
sal_mod <- lm(salary ~ yrs.service + sex, data = Salaries)
```
```{r, eval=FALSE}
get_regression_table(sal_mod)
```
```{r, echo=FALSE}
get_regression_table(sal_mod) %>% 
  kable(format = 'html')
```

$$\hat{Salary}=92357+748 \times YrsEmployed+9072 \times Sex$$

- Note that Female is the reference group, Sex = 1 for Male and 0 for Female

---
class: inverse, middle, center

# We aren't required to model the slopes to be parallel. Could allow the effect of years employed to differ between male and female professors.

---
# Visualizing Interaction Model

.pull-left[
```{r, eval=FALSE}
ggplot(Salaries, 
  aes(x = yrs.service, 
  y = salary, 
  color = sex)) + #<<
  geom_point(alpha = 0.6, 
             size = 5) +
  geom_smooth(method = 'lm', 
              se = FALSE, 
              linewidth = 2) +
  theme_minimal() +
  theme(text = element_text(size = 20))
```
]

.pull-right[
```{r, echo=FALSE, message=FALSE, out.width='99%'}
ggplot(Salaries, aes(x = yrs.service, y = salary, color = sex)) +
  geom_point(alpha = 0.6, size = 5) +
  geom_smooth(method = 'lm', se = FALSE, linewidth = 2) +
  theme_minimal() +
  theme(text = element_text(size = 20))
```
]

---
# Interaction model implication

$$y = \beta_0 + \beta_1x + \beta_2d + \beta_3 xd + \epsilon$$
$$\hat{y} = b_0 + b_1x + b_2d + b_3 xd$$
- Now $x$ and $d$ are interacted (multiplied together), allowing the effect of $x$ to differ between groups of $d$

---
# Interaction model equations

$$\hat{y} = b_0 + b_1x + b_2d + b_3 xd$$
--

- For the $d=0$ group:

$$\hat{y} = b_0 + b_1x$$

--

- For the $d=1$ group:

$$\hat{y} = b_0 + b_1x + b_2 + b_3x$$

$$\hat{y} = (b_0 + b_2) + (b_1+b_3)x$$

- Intercept *and* slope *can* differ between groups

- Intercept differs by $b_2$ and slope (affect of $x$) differs by $b_3$


---
# Interaction

- An interaction allows the effect of one variable to differ depending on the value of another variable

$$Salary=\beta_0+\beta_1YrsEmp+\beta_2Sex + \beta_3YrsEmp*Sex + \epsilon$$

- $\beta_3$ tells us how the slopes compare
  - For Sex = 1 (male), the association between years and salary is $\beta_1+\beta_3$
  - For Sex = 0 (female), it is $\beta_1$

- $\beta_2$ tells us how the intercepts compare
  - For Sex = 1 (male), intercept is $\beta_0+\beta_2$
  - For Sex = 0 (female), it is $\beta_0$

---
# Running an interaction model

```{r}
sal_mod2 <- lm(salary ~ yrs.service + sex + 
                 yrs.service*sex, #<<
               data = Salaries)
```
```{r, eval=FALSE}
get_regression_table(sal_mod2)
```
```{r, echo=FALSE}
get_regression_table(sal_mod2) %>% 
  kable(format = 'html', digits = 2)
```

$$\hat{Sal}=82069+1637*YrsEmp+20129*Sex - 932*YrsEmp*Sex$$

---
# Interpreting an interaction model

$$\hat{Sal}=82069+1637*YrsEmp+20129*Sex - 932*YrsEmp*Sex$$
- Recall that `Sex` = 1 for male, 0 for female

- What is the predicted change in salary for a male professor given one more year of employment?

- What is the predicted change in salary for a female professor given one more year of employment?

---
# Interpreting an interaction model

$$\hat{Sal}=82069+1637*YrsEmp+20129*Sex - 932*YrsEmp*Sex$$

- What is the predicted salary for a female professor employed for 10 years?

- What is the predicted salary for a male professor employed for 10 years?
