---
title: "PADP 7120 Data Applications in PA"
subtitle: "Regression with Categorical Variables"
author: "Alex Combs"
institute: "UGA | SPIA | PADP"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  xaringan::moon_reader:
    css: 'mypres.css'
    nature:
      beforeInit: "macros.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
# Outline

```{r, include=FALSE}
library(tidyverse)
library(moderndive)
library(knitr)
library(carData)
options(scipen = 999)
```

- Add categorical variables as explanatory variables

- Use a binary categorical variable as an outcome

---
# Adding to our toolbox

- So far, we have covered simple and multiple regression using quantitative/numerical variables

.font120[$$y=\beta_0+\beta_1x_1+\beta_2x_2+\dots+\beta_kx_k+\epsilon$$]

- Any number of explanatory variables could be categorical

- The outcome could be categorical as well

---
# Example

- Suppose HR wants to examine if male professors are paid differently than female professors (later coverage of inference will address whether any difference is statistically significant)

```{r}
glimpse(Salaries)
```

- discipline coded as "A" for theoretical and "B" for applied

---
# Example

.pull-left[
```{r, eval=FALSE}
ggplot(Salaries, 
       aes(x = yrs.service, 
           y = salary)) +
  geom_point(size = 4) +
  theme_minimal() +
  theme(
    text = element_text(size = 20))
```
]

.pull-right[
```{r, echo=FALSE, out.width='99%'}
ggplot(Salaries, aes(x = yrs.service, y = salary)) +
  geom_point(size = 4) +
  theme_minimal() +
  theme(text = element_text(size = 20))
```
]

---
# Example

.pull-left[
```{r, eval=FALSE}
ggplot(Salaries, 
       aes(x = yrs.service, 
           y = salary, 
           color = sex)) + #<<
  geom_point(size = 4,
             alpha = 0.5) +
  theme_minimal() +
  theme(text = element_text(size = 20))
```
]

.pull-right[
```{r, echo=FALSE, out.width='99%'}
ggplot(Salaries, aes(x = yrs.service, y = salary, color = sex)) +
  geom_point(size = 4,
             alpha = 0.5) +
  theme_minimal() +
  theme(text = element_text(size = 20))
```
]

---
# Visualizing Parallel Slopes

```{r, echo=FALSE, fig.width=8, fig.align='center'}
ggplot(Salaries, aes(x = yrs.service, y = salary, color = sex)) +
  geom_point(size = 4, alpha = 0.5) +
  geom_abline(slope = 747, intercept = 92357, color = 'red', linewidth = 2) +
  geom_abline(slope = 747, intercept = 101429, color = 'blue', linewidth = 2) +
  theme_minimal() +
  theme(text = element_text(size = 20))
```

---
# Parallel Slopes Model Implication

$$y = \beta_0 + \beta_1x + \beta_2 d + \epsilon$$
$$\hat{y} = b_0 + b_1x + b_2 d$$
- where $x$ is continuous and $d$ is a dummy variable = 0 or 1

- $d$ allows the intercept to differ between groups, but assumes affect of $x$ is the same for both

- $d$ = 0 is the reference group to which $d$ = 1 is compared

---
# Parallel Slopes Model Equations

$$\hat{y} = b_0 + b_1x + b_2 d$$

- $b_1$ is still the predicted change in $y$ due to a one-unit increase in $x$, regardless of $d$

--

- In cases where $d=0$

$$\hat{y} = b_0 + b_1x$$

--

- In cases where $d=1$

$$\hat{y} = b_0 + b_1x + b_2$$
$$\hat{y} = (b_0 + b_2) + b_1x$$
--

- $b_2$ shifts the intercept. Regression line for $d=1$ group will be above or below $d=0$ group by $b_2$ units.

---
# Example: Parallel Slopes

$$Salary=\beta_0+\beta_1YrsEmployed+\beta_2Sex + \epsilon$$
```{r}
sal_mod <- lm(salary ~ yrs.service + sex, data = Salaries)
```
```{r, eval=FALSE}
get_regression_table(sal_mod)
```
```{r, echo=FALSE}
get_regression_table(sal_mod) %>% 
  kable(format = 'html')
```

$$\hat{Salary}=92357+748 \times YrsEmployed+9072 \times Sex$$

- Note that Female is the reference group, Sex = 1 for Male and 0 for Female

---
# Interpreting results

$$\hat{Salary}=92357+748 \times YrsEmployed+9072 \times Sex$$

- What is the standard interpretation for the `YrsEmployed` coefficient?

--

- What is the standard interpretation of the coefficient for `Sex`? Recall that Sex=1 for Male and 0 for Female.

---
# Interpreting results

$$\hat{Salary}=92357+748 \times YrsEmployed+9072 \times Sex$$

- What is the predicted change in salary for a male professor following three additional years of employment?

--

- For a female professor?

--

- We forced the slopes to be parallel. Therefore, the predicted change must be the same across all groups.

---
# Interpreting results

$$\hat{Salary}=92357+748 \times YrsEmployed+9072 \times Sex$$

- What is the predicted salary for a male professor employed 20 years?

--

- For a female professor employed 20 years?

--

- Predicted value will **not** be the same because each group has a different intercept

---
class: inverse, middle, center

# We aren't required to model the slopes to be parallel. Could allow the effect of years employed to differ between male and female professors. We will revisit this after the midterm.


---
class: inverse, middle, center

# Linear probability model (LPM)

---
# LPM

- Regression model with a two-level (binary, dummy) categorical variable for the response

- Typically, `y=1` if an outcome did occur, `y=0` if it did not occur

- An LPM estimates the effect of our explanatory variables on the probability that the outcome occurs

$$Pr(y=1)=\beta_0+\beta_1x+\dots+\beta_kx_k+\epsilon$$

---
# Running an LPM

```{r, include=FALSE}
load("lectures_files/divorce.RData")
divorce <- rename(divorce, Age25to29 = Age)
```

```{r, echo=FALSE}
head(divorce, n=5) %>% 
  kable(format = 'html')
```

- Suppose we want to know how these variables might affect the probability of divorce among married or previously married individuals (Income is in $1,000s)

- Divorce = 1 if individual is divorced

- Income in $1,000s

---
# Running an LPM

```{r}
divorce_mod <- lm(Divorce ~ Age25to29 + Income + Children,
              data = divorce)
```

```{r, eval=FALSE}
get_regression_table(divorce_mod)
```

```{r, echo=FALSE}
get_regression_table(divorce_mod) %>% 
  kable(format = 'html')
```

$$\hat{Pr(y=1)}=0.38+0.4*Age+0.001*Income-0.18*Children$$

---
# Interpreting an LPM

$$\hat{Pr(y=1)}=0.38+0.4*Age+0.001*Income-0.18*Children$$

- How does having a child affect the probability of divorce?

--

- Probability is expressed in units of percent. Remember that a unit change in percent probability is a **percentage point** change.

- What is the probability of divorce for someone 27 years old making 50,000 and has 0 children?