---
title: "PADP 7120 Data Applications in PA"
subtitle: "RLab 7: Reporting Regression Results"
author: "Alex Combs"
institute: "UGA | SPIA | PADP"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  xaringan::moon_reader:
    css: 'mypres.css'
    nature:
      beforeInit: "macros.js"
      highlightStyle: github
      countIncrementalSlides: false
      highlightLines: true
---
# Outline

- Report regression results using more common tables and graphs

- This lab involves a lot of copying-and-pasting of code. You want to have these slides open on your computer. 

```{r, include=FALSE}
library(tidyverse)
library(moderndive)
library(knitr)
library(fivethirtyeight)
#library(carData)
library(stargazer) # for regression tables
library(jtools) # for regression visualizations
library(interactions) # for regression viz of interactions
#library(rockchalk)
```

---
# Setup

> **Start a new project and Rmd**

> **Change YAML.**

```{r, eval=FALSE}
---
title: "RLab 7: Reporting Regression"
author: "Your Name"
output: 
  html_document:
    theme: spacelab
    df_print: paged
---
```

---
# Setup

> **In setup code chunk, load following packages**

```{r, eval=FALSE}
library(tidyverse)
library(moderndive)
library(fivethirtyeight)
library(stargazer) # for better regression tables
library(jtools) # for better regression visualizations
library(interactions) # for better regression visualizations
```

---
# Setup

> **Add below code to setup code chunk**

```{r, warning=FALSE, message=FALSE}
state_data <- hate_crimes %>%
  filter(state != "District of Columbia") %>%
  mutate(urbanization = if_else(share_pop_metro<=0.75, "low", "high")) %>% 
  select(state, 
         median_house_inc, 
         share_pop_hs,
         share_unemp_seas,
         urbanization) %>%
  mutate(share_pop_hs = share_pop_hs*100,
         share_unemp_seas = share_unemp_seas*100)
```

---
# Regression Model

$$MedInc = \beta_0 + \beta_1 \%HS + \beta_2Urbanization + \beta_3Unemp + \epsilon$$

> **In a new code chunk, add code that runs the above regression model.**

```{r, include=FALSE}
inc_mod <- lm(median_house_inc ~ share_pop_hs + urbanization + share_unemp_seas, data = state_data)
```

---
# Reporting Results

- We have covered how to produce a standard table of results using `get_regression_table`

```{r, echo=FALSE}
get_regression_table(inc_mod) %>% 
  kable()
```

- This table is okay but not typically how results are reported

---
# Regression Tables

- Regression results tables usually include:
  - estimate
  - standard error
  - asterisks (stars) for statistical significance based on p-value
  - clear variable names
  - number of observations

- Sometimes goodness-of-fit measure like $R^2$  
- Sometimes include confidence intervals instead of standard error

---
# Regression Tables

> **In a new code chunk, add below code. Include the code chunk option `results='asis'`. Knit to HTML.**

```{r, eval=FALSE}
stargazer(inc_mod,
          type="html",
          title="Regression Results",
          dep.var.labels=c("Median Household Income"),
          covariate.labels=c("Percent High School", 
                             "Low Urbanization",
                             "Unemployment Rate"),
          keep.stat=c("n"), # reports number of observations
          digits = 0, # sets the number of decimals to report
          digits.extra = 3) # number of decimals if rounding results in 0
```

---
# Better Regression Tables

> **Change options to include confidence intervals instead of standard errors and include goodness-of-fit. Knit to HTML.**

```{r, eval=FALSE}
stargazer(inc_mod, 
          ci=TRUE, ci.separator = " : ", #<<
          type="html",
          title="Regression Results",
          dep.var.labels=c("Median Household Income"),
          covariate.labels=c("Percent High School", 
                             "Low Urbanization",
                             "Unemployment Rate"),
          keep.stat=c("n","rsq","ser"), #<<
          digits = 0,
          digits.extra = 3)
```

---
# Better Regression Tables

> **Add the last line below. Knit to HTML.**

```{r, eval=FALSE}
stargazer(inc_mod, ci=TRUE, ci.separator = " : ",
          type="html",
          title="Regression Results",
          dep.var.labels=c("Median Household Income"),
          covariate.labels=c("Percent High School", 
                             "Low Urbanization",
                             "Unemployment Rate"),
          keep.stat=c("n","rsq","ser"),
          digits = 0,
          digits.extra = 3,
          out = "inc_mod.htm") #<<
```

- This will generate a new document named `inc_mod.htm` that you can open in Word.

---
class: inverse, center, middle

# We may want to use graphs instead of tables.

---
# Visualizing Regression

- One option is to use `ggplot`

.pull-left[
```{r, eval=FALSE}
ggplot(state_data, 
       aes(x = share_pop_hs, 
           y = median_house_inc, 
           color = urbanization)) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE) +
  theme_minimal() +
  theme(title = element_text(size = 16),
        axis.text = element_text(size = 12))
```
]

.pull-right[
```{r, echo=FALSE, message=FALSE, out.width='99%'}
ggplot(state_data, 
       aes(x = share_pop_hs, 
           y = median_house_inc, 
           color = urbanization)) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE) +
  theme_minimal() +
  theme(title = element_text(size = 16),
        axis.text = element_text(size = 12))
```
]

---
# Visualizing Regression Line

- `ggplot` can easily visualize regression models that include:

  - **ONE** numerical explanatory variable
  - **ONE** categorical explanatory variable (with or without interaction)

--

- `ggplot` cannot show us the relationship between education and income between high and low urbanization controlling for unemployment rate

---
# Visualizing Regression (effect_plot)

> **Add below code in a new code chunk**

```{r, fig.align='center', fig.height=4, fig.show='hide'}
effect_plot(inc_mod, # Saved reg results
            pred = share_pop_hs, # an explanatory variable
            plot.points = TRUE, # show points or not
            data = state_data, # data used for regression
            y.label = "Median household income", 
            x.label = "% High School or Above")
```

---
# Visualizing Regression (interact_plot)

- To include urbanization, can use `interact_plot` from `interactions` package 

> **In a new code chunk, add below code**

```{r, fig.align='center', fig.height=4, warning=FALSE, fig.show='hide'}
interact_plot(inc_mod, 
              pred = share_pop_hs, 
              modx = urbanization, # the interaction variable
              plot.points = TRUE, 
              data = state_data,
            y.label = "Median household income", x.label = "% High School or Above")
```

---
# Visualizing Coefficients

- May want to visualize the estimates instead of using a table. 

> **Add following code in a new code chunk.**

```{r, warning=FALSE, message=FALSE, error=FALSE, fig.height=4}
plot_coefs(inc_mod)
```

---
# Wrap up

> **Knit to HTML. Let's examine what all this code did.**

> **Upload your Rmd to eLC**

