---
title: "PADP 7120 Data Applications in PA"
subtitle: "Lab 1: Intro to R"
author: "Alex Combs"
institute: "UGA | SPIA | PADP"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  xaringan::moon_reader:
    css: 'mypres.css'
    nature:
      beforeInit: "macros.js"
      highlightStyle: github
      countIncrementalSlides: false
---
class: inverse, middle, center

# Does everyone have access to RStudio?

---
# Objectives

- Open and navigate RStudio
- Create an R project to organize our work
- Load R packages
- Start an R markdown document and run code
- Export your R markdown document to HTML

---
# How to Follow Lab Slides

- Bullets provide background, concepts, etc.

- Block quotes in bold like below contain a prompt

> **Do this thing in this block**

- Example R code is in blocks like below

```{r, eval=FALSE}
summary(cars)
```

- Or `in-line code like this`

---
# Lab Process

- I will complete Labs along with you to help show what to do

- Helps to have the slides open on your computer

- You can copy-paste code from slides to your RStudio

---
# Lab Process

- Highly likely that you will encounter errors. This is normal. This lab will deliberately cause a few errors.

  - If so and you cannot figure out how to fix, let me know
  - Troubleshooting can help you learn

- First few weeks may involve frequent stopping as I assist students.

- If you are waiting while I assist others, your patience is appreciated. Try to resist working ahead. Maybe help a neighbor.

- Possible I can't solve an issue during class and need to resolve after class.

---
# Starting RStudio

> **Launch RStudio**

- Should see 3 Panes
  - Console (left)
  - Environment (top-right)
  - Files and other tabs (bottom-right)

---
# Working Directory

- Working directory is the folder RStudio is currently using to import and export files

> **Type `getwd()` into the console (bottom-left pane) and hit Enter**

- R should return a file path, such as "/Home/Documents/"

- When you tell R to look for or save a file, this is the location R is using

---
# RStudio Projects

- RStudio Projects are a good way to organize your work

- Creates a folder for RStudio and sets it as the working directory

- Places a `.Rproj` in the folder you can open to return to your project

- All of our Labs will begin with this process

--

> **Create a project called "Lab1" in a location of your choice (I will demonstrate)**

--

---
# Changing settings

- If using RStudio locally on your own computer, let's change some settings that should reduce the chance of errors

> **Go to "Tools" in the RStudio menu bar at the top. Click on "Global Options" at the bottom of the menu.**

> **Uncheck all boxes above "Other" and change the "Save workspace to .RData on exit" to Never.** 

---
# Loading Packages

- Many functions in R require us to load a package

- All assignments will include a list of packages you need to complete the work

- The following code loads a package

```{r, eval=FALSE}
library(name_of_package)
```

- We almost always need to load a package named `tidyverse`

> **Run the following code in your console**

```{r, eval=FALSE}
library(tidyverse)
```

---
# Installing packages

- You must have a package installed in your version of RStudio in order to load it

- If RStudio failed to load tidyverse, likely due to not having it installed (or some other package it requires). Error message will tell you a package "does not exist"

- The following code will install a package

- Make sure you include quotes around the name

```{r, eval=FALSE}
install.packages("name_of_package")
```

---
# Loading vs. Installing packages

- Every time you open RStudio, you need to re-load the packages you need

- You only need to install a package once per version of RStudio

- Only install packages using the console. Never install packages inside the R Markdown documents we will use for assignments.

---
# Importing CSV data

> **Download the file titled "states1.csv" from eLC. Leave it wherever your computer saved it (probably Downloads folder).**

--

> **Provided tidyverse loaded successfully, run the following code in your console:**

```{r, eval=FALSE}
states1 <- read_csv("states1.csv")
```

- You must use quotes anytime you tell R to load or save a file

---
# A common error

- Should now see an error message like the one below

```{r, eval=FALSE}
Error in load(states1.RData) : object 'states1.csv' not found
Calls: <Anonymous> ... withCallingHandlers -> withVisible -> eval -> eval -> load
Execution halted
```

- R tries to tell you what went wrong in the error message. This says file (object) cannot be found.

---
# Moving a file to our project

- Remember, R looks for files in your project folder or your current working directory (`getwd()`) if not working in a project

- R is saying it can't find the file we just told it to import because it is in our downloads folder, not our project folder

--

> **Using your computer's file manager, move states1.csv to your project folder**

> **Try to run `states1 <- read_csv("states1.csv")` again in your console**

---
# Data in your R session

- The environment pane (top-right) will list all data currently loaded

- You should see `states1` in your environment pane

- You can click the dropdown arrow to see a list of variables

- Or click on `states1` to see the data in spreadsheet format

---
# Starting an R Markdown Document

- R Markdown: allows you to fluidly combine content and code to export to all sorts of things [video](https://rmarkdown.rstudio.com/lesson-1.html)

- We will use R Markdown documents exclusively in this course

--

> **Start an R Markdown (.Rmd) document. When the dialog box appears, click OK.**

- By default, an Rmd loads with a template. The template information is informative for beginners.

---
# Knit

- Knit refers to taking the combination of text and code and knitting it into a HTML, PDF, Word doc, slides, etc.

- Should see a knit button at the top of the editor pane. Clicking it will knit your document.

- Every time you knit, R saves your Rmd file
  
--

> **Knit your Rmd document. R will ask you to name it first. Name it "test".**

---
# YAML header

- The YAML header tells R how to format the output

> **Change the YAML to the following:**

```{r, eval=FALSE}
---
title:  'Test'
author: 'Your Name'
output: html_document
---
```

> **Knit to HTML again**

---
# Code Chunks

- Code chunks separate text from R code

- To add a code chunk, you can use point-and-click or the keyboard shortcut `Cmd+Opt+I` on Mac, `Ctrl+Alt+I` on Windows

--

> **At the bottom of your Rmd, insert a code chunk and add some arbitrary math inside it.**

> **Click `Enter` a few times and type another line of math inside the same code chunk**

---
# Running Code Chunks

- Several options to run code in an Rmd

- To run a particular code chunk, click on the play button. This runs ALL of the lines of code in the chunk.

- To run a single line of code, place your cursor on that line and click `Cmd+Enter` on Mac or `Ctrl+Enter` on Windows

- To run multiple code chunks, use the "Run" menu at the top of the editor pane

---
# Creating an object in environment pane

- Anytime you use the **assignment operator,** `<-`, the result will be saved in your environment pane. You already saved `states1` in your environment using `<-`.

--

> **Insert another code chunk. Add and run the following code:**

```{r, eval=FALSE}
states2 <- select(states1, state, pop_1990)
```

- Now we have two datasets. We can use states2 for whatever we needed it for while preserving states1 if we need that too.

--

> **Knit your Rmd again. (this should cause an error)**

---
# Another error

- R cannot find function `select` when knitting Rmd. But it worked when we ran the single line of code.

--

- When you knit, R runs the code in your Rmd from beginning-to-end and ignores anything you did in the console to load a package or import data. Your Rmd needs to be self-contained.

- If a code chunk in your document depends on something you did in the console, it won't knit unless we add that code to the Rmd and in the correct order.

--

> **In the first code chunk of your Rmd, add a new line of code: `library(tidyverse)`. Try to knit again.**

---
# Another error

- R can't find object `states1`, but `states1` is in our environment.

--

- We imported `states1` using the console. It is in our environment. We can use it.

- But there is no code in our Rmd document that imports `states1`. When we knit, we do not have `states1` imported

--

> **Add `states1 <- read_csv("states1.csv")` in the first code chunk BELOW loading tidyverse (because read_csv depends on tidyverse). Try to knit again.**

---
# Returning to a project

> **Since you just knitted, everything is saved. Close RStudio.**

> **To re-open your project, find the .Rproj file in your file manager and open it.** 

---
class: inverse, middle, center

# Let's create our first Lab Rmd to upload to eLC

---
# Set-up

> **Download the file "Lab1_profile" from eLC and add to your project folder**

--

> **Rename the file "firstname_lastname_profile" which you can do via your file manager (as with any file) or inside RStudio (I will demonstrate).**

--

> **In your file pane, click on the file and it will open in the top-left pane.**

> **Complete the instructions until you reach the section titled "Some of my R skills"**

---
# Some of my R skills

- Under "I know how to include and execute code chunks."

> **Add a code chunk** 

> **Inside the code chunk, add some arbitrary arithmetic**

> **Save the result of the arithmetic as an object named `math` using the assignment operator**

---
# Some of my R skills

- Under "I know how to load an R package."

> **Add another code chunk**

> **Write code that will load the tidyverse package**

---
# Some of my R skills

- Under "I know how to import a data set."

> **Add another code chunk**

> **Write code that will import states1.csv**

---
# Some of my R skills

- Under "I know one way to print a data set."

> **Add another code chunk and add the following code**

```{r, eval=FALSE}
states1
```

> **Knit your Rmd.**

---
# Upload to eLC

> **Upload your Rmd file to the Lab1 assignment dropbox on eLC.**
